---
title: "Data preparation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{data_preparation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette introduces the central data classes in *ConFluxPro* and shows how to combine all necessary data into a single `cfp_dat()` object. We start by explaining the concept of a 'profile' and how soil data is structured in the package. We then introduce three essential classes `cfp_gasdata()`, `cfp_soilphys()` and `cfp_layers_map()` that frame different data types within *ConFluxPro*. Finally, we combine these objects into a single `cfp_dat()` object that holds all information needed to start modeling soil gas fluxes.

```{r setup, message=FALSE}
library(ConFluxPro)
```

# Soil data and profiles
## What is a soil profile?
A profile in *ConFluxPro* is a distinct observation of the vertical profile of a soil. For example, this can mean the concentration profile of CO~2~ at a specific time point, or the vertical distribution of soil moisture. Profiles are identified by their `id_cols`, column names that together uniquely identify each profile. This can be whatever combination of columns happened to be in your data, but will typically include a datetime column to identify the time point or a column that identifies the site if you measured at multiple locations. Consider the `soilphys` dataset included in the package.

```{r profiles_start, message = FALSE}
library(dplyr)
data("soilphys", package = "ConFluxPro")
soilphys %>%
  relocate(site, Date, gas) %>%
  head()
```

This dataset contains information on different soil physical parameters. We will come back to that later. For now, notice that there are the `site` and `Date` column. All combinations identify a unique profile, namely twelve dates for two sites `"site_a"` and `"site_b"` for a total of 24 profiles. We can create a `cfp_profile()` object from this dataset. The only thing this does is to add the information which columns identify the profiles to the dataset with the `id_cols` argument. Internally, the number of distinct profiles is calculated as the unique combinations of the `id_cols` and can be accessed by `n_profiles()`.
```{r profiles_class}
soilphys <- 
  cfp_profile(soilphys, id_cols = c("site", "Date"))
n_profiles(soilphys)
```

The concept of profiles and `id_cols` applies to almost all objects created in ConFluxPro. You can always see what `id_cols` an object has by calling the function `cfp_id_cols()`.
```{r profiles_id_cols}
cfp_id_cols(soilphys)
```

## Soil data structure
Within a soil profile, parameters change with depth. In ConFluxPro we differentiate between two types of data. _Point data_ have distinct values at a precise `depth` in the soil, such as the soil gas concentration. For _layered data_ , the soil is subdivided into layers that are delimited by their `upper` and `lower` boundary. Layered data must be without gaps or overlaps, so that the entire profile can be described by distinct values. `depth`, `upper` and `lower` are in centimeters and a higher value indicates the upward direction. Both positive and negative values are allowed. This means you could set the mineral/organic soil boundary to have `depth = 0`, describe the mineral soil with increasingly negative values of `depth` and have the organic layer on top have positive values.

We will now present the three main data types that are needed to model gas fluxes.

# `cfp_gasdata()`: Gas concentration data
Soil gas concentration data is stored in a `data.frame`. The data needs to have at least the columns `gas`, `depth` and `x_ppm`, as well as any `id_cols` that identify the soil profiles. The `gas` is a `character()` that identifies which gas was measured, so for example `"CO2"`. `depth` is as explained above and `x_ppm` is the mixing ratio in ppm. Consider the example data that ships with the package:
```{r gasdata-example}
data("gasdata", package = "ConFluxPro")
head(gasdata)
```

Because this `data.frame` already follows the requirements defined above, we can create a `cfp_gasdata()` object right away by providing the necessary `id_cols`. The function checks, whether the `data.frame` meets the criteria and returns an object of class `cfp_gasdata`.
```{r cfp_gasdata, error=TRUE}
# create working cfp_gasdata object
my_gasdata <- cfp_gasdata(gasdata, id_cols = c("site", "Date"))

head(my_gasdata)

# won't work, because depth is missing
gasdata_broken <- gasdata %>% select(!depth)
cfp_gasdata(gasdata_broken, id_cols = c("site", "Date"))
```

The data is in a long format, following the [tidy data principles](https://tidyr.tidyverse.org/articles/tidy-data.html), where each observation is in a new row. If your data is not in this format, you can use the `tidyr::pivot_longer()` and other functions in the `tidyr` package. 
Notice that `cfp_gasdata()` automatically added the `gas` column to the `id_cols`. 
`gas` is always a required column in ConFluxPro, so that many functions will automatically add it to the `id_cols` if its present and will fail if its missing. 

# `cfp_soilphys()`: Soil physical data
## Create a `cfp_soilphys()` object
Soil physical data includes everything else needed to describe the gas transport characteristics of the soil. 
This data is split into homogeneous layers with constant values for each layer and profile.
At the minimum, `cfp_soilphys()` requires two parameters: `DS` is the gas-specific apparent diffusion coefficient of the soil in m^2^ s^-1^ and `c_air` is the number density of the soil air in mol m^-3^. 
If these parameters are provided, a valid `cfp_soilphys` object can be created. 
This function ensures that each profile is correctly layered without overlaps or gaps and that the required columns are present.

```{r soilphys-example}
data("soilphys", package = "ConFluxPro")

my_soilphys <- cfp_soilphys(soilphys, id_cols = c("site", "Date"))
```

## Derive parameters with `complete_soilphys()`
Often `DS` and `c_air` are derived from other parameters that are easier to measure. 
This then requires a bit more work to get a working `cfp_soilphys()` object.
Typically, you will need the following information: 

**Measured parameters:**

- `TPS`: total porosity as a volume fraction.
- `SWC`: soil water content as a volume fraction.
- `t`: air temperature in °C.
- `p`: air pressure in hPa.

**Derived parameters:**

- **`c_air`: air number density, derived from `t` and `p` with ideal gas law.**
- `AFPS`: air-filled pore space, defined as `AFPS = TPS - SWC`.
- `DSD0`: relative diffusivity of the soil, unitless. Derived from `AFPS` with transfer function.
- `D0`: gas-specific free-air diffusion coefficient in m^2^ s^-1^. Derived from `t` and `p` with empiric function.
- **`DS`: gas-specific apparent diffusion coefficient in m^2^ s^-1^, derived as `DS = D0 * DSD0`.**

These relationships are invoked within the `complete_soilphys()` function. Consider the provided example data: 
It already has every parameter we need. 
Now let's remove the derived parameters and run `complete_soilphys()` again. 
We have to provide a formula for the calculation of the relative diffusivity `DSD0`. 
In our case, we have fitting parameters `a` and `b` from a Currie-style model that calculates as `DSD0 = a * AFPS ^ b`. 
We provide this formula as a `character()` to the function:
```{r soilphys-complete-sp}
soilphys_measured_only <- soilphys %>% select(!all_of(c("AFPS", "DSD0", "D0", "DS", "c_air")))

head(soilphys_measured_only)

my_soilphys_completed <- complete_soilphys(soilphys_measured_only, DSD0_formula = "a*AFPS^b")
head(my_soilphys_completed)
```

Notice the `gas` column. This is essential because different gases have specific diffusion coefficients. 
We can also remove the `gas` column and provide a list of gases to `complete_soilphys()`:
```{r soilphys-complete-gases}
soilphys_measured_only %>%
  select(!gas) %>%
  complete_soilphys(DSD0_formula = "a*AFPS^b", gases = c("CO2", "CH4")) %>%
  head()
```

This replicates each row of the `data.frame` and calculates distinct values for each gas.

## From measurements to layered data: `discretize_depth()`
As seen above, a `cfp_soilphys` object often requires the combination of many different parameters.
These parameters are typically measured in very different ways and may be present in different data formats. 
This presents a hurdle, as `cfp_soilphys()` requires these datasets to all be present in the same layered format. 
The function `discretize_depth()` helps with this by providing a unified way to interpolate different observations to the same layered structure. 
The basic idea is that you provide a vector of depths that represent the layer boundaries. 
For $n$ layers, you would need $n+1$ depths. 
Let's say we measured a temperature profile at three depths: 0, -30 and -100 cm:
```{r temperature-example}
temperature_profile <- 
  data.frame(depth = c(0, -30, -100),
             t = c(25, 20, 16))
```
If we want to map this to 10 cm-thick layers, we can simply define the layer boundaries and use `discretize_depth()` to linearly interpolate the observations:
```{r temperature-interp}
boundaries <- seq(0, -100, by = -10)

discretize_depth(temperature_profile,
                 param = "t", 
                 method = "linear",
                 depth_target = boundaries)
```

The function can handle proile data and treats every profile distinctly.
So if we adapt our example data to two profiles identified by a `date` column, the interpolation works in the same way, if we provide the necessary `id_cols`.
```{r}
temperature_profile_dates <-
  data.frame(date_id = rep(c("date_1", "date_2"), each = 3),
             depth = rep(c(0, -30, -100), times = 2),
             t = c(25, 20, 16, 15, 10, 8)) %>%
  cfp_profile(id_cols = c("date_id"))

discretize_depth(temperature_profile_dates,
                 param = "t", 
                 method = "linear",
                 depth_target = boundaries)
```

We can use different interpolation methods that may be more applicable to other parameters.
See the function documentation `?discretize_depth` for details.

# `cfp_layers_map()`: Create the model layers
The final aspect of a ConFluxPro model is the "layers_map".
This is a layered `data.frame` that defines for which layers we want to calculate flux rates.
We provide a dataset that matches the other example datasets:
```{r layers_map-load}
data("layers_map", package = "ConFluxPro")

layers_map
```
As you can see, we divide each of the site into two layers, one for the organic layer with `depth > 0` and one for the mineral soil below. 
Notice, that the upper boundary is different between the sites. 
This is because they have a different organic layer, that is also reflected in the other datasets:
```{r may_upper}
my_gasdata %>%
  group_by(site) %>%
  summarise(max_depth = max(depth))

my_soilphys %>%
  group_by(site) %>%
  summarise(max_upper = max(upper))
```

Notice, that the extends of the layers_map should match your measured data if you want to include it in the models. 
You cannot extend the layers_map beyond your measurements.

How many layers to define in the `cfp_layers_map()` is up to the user and depends on the research question and the resolution of the recorded data.
While it may be tempting to define a layer between every observation depth in `cfp_gasdata()`, this may result in artefacts in the modeled fluxes.
Combining these layers may therefore yield more robust results.

To create a valid `cfp_layers_map`, we still need to add some information.
These are only needed for a `pro_flux()` inverse model and are set to `NA` if missing.

- `gas`: The gas as a `character()`.
- `lowlim`: The lower limit of production allowed in that layer in µmol m^-3^ s^-1^ 
- `highlim`: The upper limit of production allowed in that layer in µmol m^-3^ s^-1^ 
- `layer_couple`: Penalty factor, defaults to 0. See `?cfp_layers_map()`

We can either do so in the function call or provide distinct values ourselves.
The `highlim` and `lowlim` values can help to explicitly exclude unrealisitc processes in the optimization in `pro_flux()`.
For example, we can set `lowlim` to 0, if we want to prevent consumption of CO~2~.
Again, these values are only relevant in the `pro_flux()` model. 
In our case, we want to add the limits for the gas `"CO2"`:
```{r}
my_layers_map <- 
  cfp_layers_map(
    layers_map,
    id_cols = "site",
    gas = "CO2",
    lowlim = 0,
    highlim = 1000
  )

my_layers_map
```

Note that we only added `"site"` as `id_cols`, because we don't need the model frame to change over time and therefore use the same mapping for every `Date`.
You can always use a subset of the `id_cols` in `cfp_soilphys()` in your `cfp_layers_map()`.
At the minimum, you need to provide `gas` within your `id_cols`. 

With this complete, we now have everything to create a unified dataset for a ConFluxPro model.

# `cfp_dat()`: Combining the datasets
As a last step, we combine the three datasets into one unified frame with `cfp_dat()`.

```{r}
my_dat <- cfp_dat(my_gasdata, my_soilphys, my_layers_map)

my_dat
```
This function does multiple things. 
(I) It ensures that the upper and lower boundaries between the datasets match. 
The highest `upper` of `soilphys` and `gasdata` should match the highest `depth` in `gasdata`, with the `lower` boundary correspondingly.
(II) It matches each profile to another and makes a list of all profiles and the corresponding datasets.
ConFluxPro does not require the `id_cols` between the different datasets to match exactly, as long as the different profiles can be mapped to another.
This makes it possible, for example, to map the same `soilphys` profiles to multiple distinct `gasdata` profiles, without having to replicate the `soilphys` dataset manually.
We already make use of this by providing only one `layers_map` per `site` in our example that is not replicated for the different `Date` of the the other datasets.
(III) For each profile, it separates the layers of `soilphys`, at each depth where there is an observation in `gasdata`. This does not change the calculation but is needed internally for the `pro_flux()` model. 
(IV) It adds the column `pmap` to the `soilphys` dataset that indicates which layer, as defined in `cfp_layers_map()` corresponds to that layer in `soilphys`.

At its core, a `cfp_dat` object is simply a list. 
We can access the different elements, including the generated `profiles` `data.frame` like any element in a list with the `$` operator. 
```{r}
my_dat$profiles %>% head()
```

Within this `data.frame` we have identifiers for each profile of the different datasets (`gd_id`, `sp_id`, `group_id`) and an identifier of each profile of the combined dataset `prof_id`.

We now are ready model. 
