---
title: "Flux calculation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{flux}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows how to model flux rates with two functions `fg_flux()` and `pro_flux()`. 
It builds on the article on `vignette("data_preparation", package = "ConFluxPro")`, so you are encouraged to first read that one.

```{r setup}
library(ConFluxPro)
```

# General workflow
We will start by importing the `cfp_dat()` object, that we created at the end of the last article `vignette("data_preparation", package = "ConFluxPro")`.
This is included in ConFluxPro as the dataset `base_dat`.
```{r base_dat}
data("base_dat", package = "ConFluxPro")
```

Note, that one of the most important settings of both models is already made at this point: the definition of the flux layers in the `layers_map` of the data.
You need to carefully decide how many layers and which boundaries to set.
This may depend on your research question and on your input data. 
You can always try out different `layers_map` by changing them in the creation of your `cfp_dat()` object.

Modeling flux rates with both functions is as easy as passing `base_dat` as the single argument:
```{r base_flux, warning=FALSE}
mod_pf <- pro_flux(base_dat)
mod_fg <- fg_flux(base_dat)
```
This will use the default settings for both functions. We will look into the fine-tuning of both in more detail later. 
For now we are only interested in the objects returned. 
`pro_flux()` returns an object of class `cfp_pfres`, while `fg_flux()` returns a `cfp_fgres` object. 
Both have the same structure. 
At their core is the input dataset `base_dat` with the model settings as additional attributes. 
The model result is stored as another `data.frame` in the list.
```{r base_list}
names(mod_pf)
names(mod_fg)
```
These `data.frame`s hold the flux rates in the different layers, and the optimized production rates of the inverse model (`pro_flux()`). 
To derive efflux rates, we don't have to interact with these internal datasets directly. Instead, we can simply call `efflux()` on the entire model result.
This will return a `data.frame` with the efflux rates (in µmol m^-2^ s^-1^) of each profile of the model.
```{r base_efflux}
efflux_pf <- efflux(mod_pf)
efflux_fg <- efflux(mod_fg)

head(efflux_pf)
```
Similarly, we can extract the total production rates for each layer within the soil.
```{r}
production_pf <- production(mod_pf)
production_fg <- production(mod_fg)

head(production_pf)
```
Here, `prod_abs` is the total production rate in the layer (in in µmol m^-2^ s^-1^), whereas `prod_rel` is the relative contribution of that layer to the total efflux.

Now that we've discussed the syntax, let's discuss the specific considerations and settings of each method. 

# `fg_flux()`: the 'classic' FGM
## Model settings
The function `fg_flux()` implements different versions of the 'classic' FGM.
With this we mean a typical 'forward' calculation of the concentration gradient and diffusion coefficient, in contrast to the inverse model discussed below. 
In truth, `fg_flux()` is a collection of models, because there exist different philosophies on how to calculate the concentration gradient and how to extrapolate the flux rates to the surface. 
The default mode `"LL"` is to fit a local linear model of `x_ppm~depth` within each layer independently. 
Other modes fit a single model to the complete concentration profile, either a linear spline (`"LS"`), or an exponential model (`"EF"` and `"DA"`). 
We can use one or more of thse methods by using the `modes` argument. 
If we choose multiple modes, we must also specify the `gases` argument. 
This allows to choose a different mode for each gas that is most applicable.
```{r flux-modes, warning=FALSE}
mod_fg_ef <- fg_flux(base_dat, modes = "EF")
mod_fg_multiple_modes <- fg_flux(base_dat, 
                                 modes = c("LL", "LS", "EF"),
                                 gases = rep("CO2", 3))
```

In addition, we also have to specify how parameters within the `soilphys` dataset are aggregated. 
If the layers specified in `layers_map` contain multiple layers in `soilphys`, the parameters have to be averaged. 
Relevant for the flux calculation are only the parameters `c_air`, the number density of the soil air and `DS`, the apparent diffusion coefficient. 
By default, `DS` is averaged as a weighted harmonic mean, while `c_air` is averaged as a weighted arithmetic mean. 
The weights in both cases are the respective layer heights. 
You can change this, or add more parameters that are averaged by setting the arguments `param` and `funs`. 
These are vectors that must match in length. Each `param` must be present as a column in `soilphys`.

```{r flux-param}
mod_fg_more_params <- 
  fg_flux(base_dat, 
          modes = "LL",
          gases = "CO2",
          param = c("c_air", "DS", "SWC", "t"),
          funs = c("arith", "harm", "arith", "harm"))
```

## Extrapolating efflux
How to derive efflux rates from the flux rates of each soil layer requires some consideration. 
Here, we implement three different methods that can be set in the call of `efflux()`. 
The simplest is simply to assume that the flux rate in the top layer is equivalent to the efflux:
```{r efflux-top}
efflux_fg_top <- efflux(mod_fg, method = "top")
```

The other two methods extrapolate to the surface. 
Both assume that the modeled flux rate is centered in the layer, so that the flux of a layer between 0 and 10 cm depth is located at 5 cm depth.
The `"lm"` model fits a linear model `flux~depth` of all layers and extrapolates it to the surface.
The `"lex"` model uses two layers to linearly extrapolate the flux rate.
These have to be selected using the `layers` argument. 
Here we set `layers = c(1, 2)` to select the top two layers for the extrapolation.
Since we only have two layers to begin with, the results of `"lm"` and `"lex"` are the same in our case.
```{r fg_ef}
efflux_fg_lm <- efflux(mod_fg, method = "lm")
efflux_mod_lex <- efflux(mod_fg, method = "lex", layers = c(1, 2))
```

## Deriving production rates
The user input needed to derive production rates from a model returned by `fg_flux()` is simple. 
Because the efflux rate is needed, we potentially have to specify how we want to calculate the efflux rate.
For this simply pass the same arguments as with `efflux()`.
If nothing is specified, the default efflux method is used (`"lm"`).

```{r fg_prod}
production_fg <- production(mod_fg)
```

The production rates of each layer are calculated as the difference of the flux in the layer below and the flux in the layer above it.
For the lowest layer, the flux below is assumed to be zero, for the highest layer the efflux is used as the upper flux.
This process is prone to artefacts stemming from the methods used in the calculation of the concentration gradient.
Production rates derived from `fg_flux()` should therefore be carefully checked for plausibility. 

# `pro_flux()`: inverse model of production profiles
## Theory and implementation
The function `pro_flux()` implements an inverse model of soil gas production profiles.
For each layer in `soilphys`, a theoretical concentration profile is modeled.
In this way, the complete concentration profile can be modeled.
$$
c(z) = - \frac{P}{2D_S} \cdot z^2 - \frac{F_0}{D_S} \cdot z + c_0
$$
The model is evaluated at each layer boundary and "from-bottom-to-top", meaning that the lowest layer is modeled first.
For each layer in `layers_map`, a production rate $P$ (µmol m^-3^ s^-1^) is algorithmicly optimized to achieve the best fit with the measured concentration profile. 
This is achieved with the `stats::optim()` function and a custom objective function.
At the lowest layer, two boundary conditions need to be met: the concentration $c_0$ and the incoming flux $F_0$ from the deep soil. 
While $c_0$ is set to the median of the measured value at that depth, $F_0$ can be either set to zero, or optimized together with the production rates $P$.
In this model, the efflux is the sum of the production (or consumption) in the entire profile.
## Model settings
As with the 'forward' model, the most important setting is to determine for which layers a production rate is calculated, which is defined in `layers_map`.
In addition, there are now more settings that can be applied within the call to `cfp_layers_map()`.
Namely, we can set a upper `highlim` and lower limit `lowlim` of the allowed production rate within each layer.
Be limiting the range in which the production rates are optimized to positive values, we can exclude consumption of the gas within the model.
This can be relevant especially for CO~2~, where negative concentration gradients are often the result of measurement uncertainty and not of significant consumption of the gas. 
For other gases, such as CH~4~ both consumption and production may be relevant processes that can also occur at different points within the same profile.

The other important setting is `zero_flux` within the call to `pro_flux()`. 
If `TRUE` (the default), the flux from below the model boundary into the lowest layer ($F_0$) is assumed to be zero.
If `FALSE`, $F_0$ is optimized alongside the production rates within the soil.
Typically, there is no advantage in the modelling of setting `zero_flux` to `FALSE`, even if the assumption that the flux is negligible is not met.
But be aware, that in this way, more production may be allocated into the lowest layer than that layer actually produced.
If you do decide to use `zero_flux = FALSE`, you can  also set `zero_limits` that sets limits for the allowed flux rates, analogous to `lowlim` and `highlim` in `cfp_layers_map()`.

```{r pf_zf}
mod_pf_with_flux <- pro_flux(base_dat, 
                             zero_flux = FALSE,
                             zero_limits = c(-1000, 1000))
```

You can access the optimized $F_0$-flux rates within the function `deepflux()`. 

```{r pf_df}
deepflux(mod_pf_with_flux) |>
  head()
```

Other settings are more experimental, like `layer_couple` within `cfp_layers_map()` which aims to penalize stark changes of production within a profile, similar to `evenness_factor` in `pro_flux()`.
We do not encourage the use of these setting, as contrasting production rates may be a sign that you defined too many production layers.

## Extracting efflux and production rates
Getting the efflux from any `pro_flux()` model is very simple by calling `efflux()` on any model result.
No additional settings are necessary.
```{r pf_ef}
EFFLUX_pf <- efflux(mod_pf)
head(EFFLUX_pf)
```

Similarly, you can call `production()` to get the optimzed production rates for each layer in `layers_map`.
```{r pf_prod}
PRODUCTION_pf <- production(mod_pf)
head(PRODUCTION_pf)
```

