---
title: "overview"
author: Valentin Gartiser
date: '2022-04-29'
output: 
  rmarkdown::html_vignette:
    number_sections: true 
    fig_width: 7
    fig_height: 4
    toc: true
    toc_depth: 3
    
vignette: >
  %\VignetteIndexEntry{overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
<style>
    img {
        border: 0;
    }
</style>

<img src="../inst/logo.svg" width="300" height="300" style ="float:right"/>

# Introduction

Hello and welcome to ConFluxPro - a toolkit for soil gas analysis. The goal of this package was to implement the gradient-flux approach for modeling soil gas fluxes based on concentration measurements. A good starting point to get an idea of the setup and ideas behind this method is our paper Maier et al. (2020).

# Package structure
```{r, echo = FALSE, fig_height = 9}
DiagrammeR::grViz(diagram = "digraph flowchart {
graph [layout = dot, rankdir = LR]


  node [style = filled color = white shape = rectangle]
  node [fillcolor = BlueViolet fontcolor = white]
  a [label = 'gasdata']
  b [label = 'soilphys']
  c [label = 'layers_map']
  h [label = 'cfp_dat']
  m [label = 'cfp_fgres']
  n [label = 'cfp_pfres']
  o [label = 'EFFLUX']
  p [label = 'PRODUCTION']
  s [label = 'cfp_altres']
  
  node [fillcolor = orange shape = oval fontcolor = black]
  d [label = 'cfp_dat()']
  e [label = 'cfp_gasdata()']
  f [label = 'cfp_soilphys()']
  g [label = 'cfp_layers_map()']
  i [label = 'fg_flux()']
  j [label = 'pro_flux()']
  k [label = 'efflux()']
  l [label = 'production()']
  q [label = 'run_map()']
  r [label = 'alternate()']
  
  a -> {e}
  b -> {f}
  c -> {g}
  {e f g} -> d
  d -> h
  h -> {i j}
  i -> m
  j -> n
  {m n} -> {k l r}
  k -> o
  l -> p
  q -> r
  r -> s
  
   subgraph clusterMain {
   graph [rankdir = TB]
   node [style = filled color = white shape = rectangle]
  node [fillcolor = BlueViolet fontcolor = white]
  aaa [label = 'objects']


  node [fillcolor = orange shape = oval fontcolor = black]
  bbb [label = 'functions()']
   
}
  
}
 
  ")
```
blablalba


# General workflow
## data preparation

## Flux-Gradient-Method FGM

## Inverse Flux moddeling pro_flux()

